Community-Centric Innovation

### 2014-02-21 paper.tex
European projects

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{he} growth of the Internet has dramatically accelerated the path of technological evolution.
The amount of information and knowledge available on the Net is increasing as well as the number of people that can access that knowledge.
Still, it is important to realize that the Internet is not simply an incremental evolution that brings faster progress.
The Internet is a disruptive transformation that is completely transforming the way we organize ourselves and relate to each other.

The fact that the information can be shared with little restrictions in terms of space, time and cost, makes it possible for people to organize in horizontal structures.
In this horizontal models, also known as peer-to-peer, all the participants can access to the information required to contribute to the decision-making process.
The heterogeneous background of the participants makes it possible to consider a wider space of solutions and the extensive scrutiny reduces the number of mistakes.

@@

Another aspect that can draw the attention to community networks is privacy.
As commercial networks are routinely involved in mass surveillance programs, the citizens see in community networks a way to opt out from privacy-invading campaigns.
Finally, community networks are also means of getting around censorship.
Totalitarian regimes include censorship into the networks to control which information is distributed among the population.
Filtering the information and ordering Internet blackouts have been used as means to hinder social movements and placate population unrest.

@@

Mesh technologies such as the \emph{quick mesh project} (qmp) are evolving to make it easier for regular citizens to set up a network in a matter of minutes.
Friendly web interfaces for configuration abstract the underlying technological complexity.
As a consequence, it is possible to deploy and maintain network connectivity even in the most adverse conditions.

The decentralized nature of community networks greatly helps in fostering organic growth and resisting node failures.
The absence of central points of failure guarantees that two neighbours will be able to connect to each other as long as they keep their nodes running, regardless of what happens to the rest of the network.
The dynamic routing protocols, such as BMX6 allows the network to quickly re-configure itself and route packets around failures to keep connectivity.
To take a community network down it is necessary to shut down all the nodes.

@@

This continuous change represent a challenge for spying agencies interested in compromising the network.
It is much more difficult to attack up-to-date system that those that run old unpatched firmwares.

@@

The participation in a community network trains the citizens in self-management techniques that inherit many of the properties of decentralized networks.
The lack of a central authority obliges the participants to take responsibility, exercise their freedom respecting others and make concessions in order to achieve common goals.
The organizational structures that we find in community networks mimic those of free (libre) software communities and other Internet-based coordinated projects such as Wikipedia.

All participants are invited to contribute and contributions are evaluated in an informal process for merit or weaknesses.
The participants can progressively build up a reputation by making high-quality contributions and helping others to accomplish their goals.
The organization progressively evolves towards a meritocratic systems in which those participants that consistently contribute to the value of the community earn the respect of their peers and emerge as natural leaders.

@@

A core principle of community networks is precisely the sharing of the network as a commons good.
There is no single owner of the network.
In fact there is no owner of the network.
Even though the participants may own the pieces of hardware that they contributed to the network, the network as a whole belongs the community.
As the value of the network increases with the number of participants, the community have an incentive to freely share network resources to invite others to join the network.

@@

Another area that has witnessed the potential for community-centric innovation is that of networked cryptocurrency.
The combination of cryptography, free libre software and a data network make it possible the appearance of networked cryptocurrencies.
In these technologies, a cryptographically secure distributed ledger is maintained by the network.
The network is composed of nodes that interact in a peer-to-peer fashion following a common set of rules or protocol to guarantee the integrity of the cryptocurrency.
The network not only distributes information about the ongoing transactions, it also provides hashing power that is used as a proof of work.
The distributed ledger is protected by the proof of work.
In a sense, the nodes vote with their proof-of-work and the majority decides which transactions are included in the public distributed ledger.

Cryptocurrency has some advantages when compared to traditional fiat currencies.
Some of the advantages of cryptocurrencies are related to their peer-to-peer nature.
One of them is the lack of a central authority that can make arbitrary decisions and create money at no cost.
In cryptocurrencies, the emission of money is controlled by the protocol and governed by publicly known rules.
This means that the availability of money at any point in the future is known by all the participants.
The emitted money is used to reward the nodes that contribute with hashing power to the security of the network.
Hashing power requires hardware and electricity, which means that is no longer possible to create money at zero cost.
In fact, the money generation business (popularly known as \emph{mining}) is a risky endeavour as the cost of hashing power can exceed the value of the generated money.

Peer-to-peer networks are extremely resilient.
There is no single point of failure or central entity that can be attacked to take down the network.
As there are thousands of nodes spread across the network, the currency can survive as long as the underlying data network survives.
The risks of blockades or political interference are greatly mitigated.
Many Internet organizations such as Wikileaks and Wordpress accept donations and payments in cryptocurrency to bypass totalitarian governments blockades or lacking payment networks.

Some of the advantages of networked cryptocurrencies are derived from their networked nature.
The scope of the cryptocurrency is the same as the underlying network.
If a cryptocurrency runs on the Internet, the scope is global.
As a result, economic transactions benefit from all the advantages of Internet protocols.
Sending money is as simple as sending an electronic mail.
It propagates almost instantaneously across all the Internet and is confirmed in a matter of minutes.
Payments can be easily automatized using scripts or other software and emitted from any networked device, such as servers, laptops or mobile phones.
Furthermore, the network is available 24 a day 365 days a year.

There are also properties of networked cryptocurrencies related to their cryptographical nature.
These currencies do not require the users to trust a third party.
The trust relies on cryptographic primitives that have proven to be robust over the years.
The reliance on secret keys makes it possible to save the same money in more than one place at the same time.
Even if one of the copies of the secret keys is lost, the owner can still access the money if she has access to a second copy.
More sophisticated techniques can be used to achieve greater degrees of security. 
For example, the secret can be divided in three parts in such a way that two out of the three parts are needed to access the money.
Secrets can also be kept in the brain, with no physical proof of the ownership of the money.

@@

The use of networked criptocurrencies evidences that flows of money are nothing more than flows of information.
Money transactions are simply communications among the involved parties.
The freedom to use different currencies is directly related to the freedom to communicate, which is popularly referred to as free speech.
The flows of money within a community are communications flows that strengthen the relation between the members of the community and therefore the community as a whole.

@@

The gift (or reputation) economy is the kind of economy we find on the Internet, in which the participants offer software, articles, music, pictures and all kind of digital goods for free using permissive licenses such as Creative Commons.
In the information age in which is obvious that money is just information and wealth is measured in knowledge instead of material goods, the progressive evolution towards a reputation economy is a natural step.

@@

Throughout human history, organizational models have changed, transformed and evolved.
As technology evolves, old models are been progressively replaced by others that more efficient in the context of the new technologies.
The invention of information processing and transmission tools favors community-centric models.
Initiatives embracing the community centric model have an advantage compared to those attaching to old hierarchical approaches.

The first example of successes can be found precisely in the fields of information and communication technologies.
The Internet itself is probably the best example of a community-centric endeavour.
There is a large number of people fascinated by and contributing to the Internet.
The decision-making forums are open and anyone is free to contribute expertise.
In a first level the users organize as networks, and then at another level those networks interconnect among them using peering agreements.

Another example of the successful application of the community-centric model is the free libre opens source movement.
Thousands of programmers affiliate to and contribute to this movement working as self-managed community.
The goods offered by this community often outperform those that have been developed using hierarchical models.
By making it possible for anyone to contribute, FLOSS projects have an advantage with respect to those that choose more restrictive models.

@@

>>>>

\section{Introduction}
% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps.
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{T}{he} growth of the Internet has dramatically accelerated the path of technological evolution.
The amount of information and knowledge available on the Net is increasing as well as the number of people that can access that knowledge.
Still, it is important to realize that the Internet is not simply an incremental evolution that brings faster progress.
The Internet is a disruptive innovation that is completely transforming the way we organize ourselves and relate to each other.

The fact that the information can be shared with little restrictions in terms of space, time and cost, makes it possible for people to organize in horizontal structures.
In this horizontal models, also known as peer-to-peer, all the participants can access to the information required to contribute to the decision-making process.
The heterogeneous background of the participants makes it possible to consider a wider space of solutions and the extensive scrutiny throughout the discussion and decision-making stage reduces the chances of mistakes.

@@

Another aspect that can draw the attention to community networks is privacy.
As commercial networks are routinely involved in mass surveillance programs, the citizens see in community networks a way to opt out from privacy-invading campaigns.
Finally, community networks are also a means of getting around censorship.
Totalitarian regimes build censorship into the networks to control which information is distributed among the population.
Filtering the information and ordering Internet blackouts have been used as ways for hindering social movements and placate population unrest.

@@

Mesh technologies such as the \emph{quick mesh project} (qmp) are evolving to make it easier for regular citizens to set up a network in a matter of minutes.
Friendly web interfaces for configuration abstract the underlying technological complexity.
In community networks local people become familiar with the procedures required to set up, maintain and troubleshoot the network.
In case of emergency, the presence of local experts facilitates quick deployments and reparations.

The decentralized nature of community networks greatly helps in fostering organic growth and resisting node failures.
The absence of central points of failure guarantees that two neighbours will be able to connect to each other as long as they keep their nodes running, regardless of what happens to the rest of the network.
The dynamic routing protocols, such as BMX6, allow the network to quickly re-configure itself and route packets around failures to keep connectivity.
To take a community network down it is necessary to shut down all the nodes.

@@

This continuous change that forces to keep the equipment firmware updated represents  an additional layer of protection against spying agencies interested in compromising the network.
It is much more difficult to attack up-to-date system that those that run old unpatched firmwares.

@@

The participation in a community network trains the citizens in self-management techniques that inherit many of the properties of decentralized networks.
The lack of a central authority obliges the participants to take responsibility, exercise their freedom respecting the others and making concessions in order to achieve common goals.
%The organizational structures that we find in community networks mimic those of free (libre) software communities and other Internet-based coordinated projects such as Wikipedia.

All participants are invited to contribute, and contributions are evaluated in an informal process for merit or weaknesses.
The participants can progressively build up a reputation by making high-quality contributions and helping others to accomplish their goals.
The organization progressively evolves towards a meritocratic systems in which those participants that consistently contribute to the value of the community earn the respect of their peers and emerge as natural leaders.

@@

A core principle of community networks is precisely the sharing of the network as a commons good.
There is no single owner of the network.
%In fact there is no owner of the network.
Even though the participants may own the pieces of hardware that they contributed to the network, the network as a whole belongs the community.
As the value of the network increases with the number of participants, the community has an incentive to share existing network resources with newcomers.

@@

Another area that has witnessed the potential for community-centric innovation is that of networked cryptocurrency.
The combination of cryptography, free libre software and a data network makes it possible the use of networked cryptocurrencies.
In these technologies, a cryptographically secure distributed ledger is maintained by the network.
The network is composed of nodes that interact in a peer-to-peer fashion following a common set of rules or protocol to guarantee the integrity of the cryptocurrency.
The network not only distributes information about the ongoing transactions, it also provides hashing power that is used as a proof of work.
The distributed ledger is protected by the proof of work.
In a sense, the nodes vote with their proof-of-work and the majority decides which transactions are included in the public distributed ledger.

Cryptocurrency has some advantages when compared to traditional fiat currencies.
Some of the advantages of cryptocurrencies are related to their peer-to-peer nature.
One of them is the lack of a central authority.
A central authority can make arbitrary decisions and create money at no cost.
In cryptocurrencies, the emission of money is controlled by the protocol and governed by publicly known rules.
This means that the availability of money at any point in the future is known by all the participants.
The emitted money is used to reward the nodes that contribute with hashing power to the security of the network.
Hashing power requires hardware and electricity, which means that is no longer possible to create money at zero cost.
In fact, the money generation business (popularly known as \emph{mining}) is, in a financial sense, a risky endeavour.
The cost of hashing power required to \emph{mine} can exceed the value of the generated money.

Peer-to-peer networks are extremely resilient.
There is no single point of failure or central entity that can be attacked to take down the network.
As there are thousands of nodes spread across the network, the currency can survive as long as the underlying data network survives.
The risks of blockades or political interference are greatly mitigated.
Many Internet organizations such as Wikileaks and Wordpress accept donations and payments in cryptocurrency.
In some cases to bypass totalitarian governments blockades and in others to operate in countries with lacking payment networks.

Some of the advantages of networked cryptocurrencies are derived from their networked nature.
The reach of the cryptocurrency is the same as that of the underlying network.
If a cryptocurrency runs on the Internet, the scope is global.
As a result, economic transactions benefit from all the advantages of Internet protocols.
Sending money is as simple as sending an electronic mail.
It propagates almost instantaneously across all the Internet and is confirmed in a matter of minutes.
Payments can be easily automatized using scripts or other software.
Any networked device, such as servers, laptops or mobile phones, can be used to send a payment.
Furthermore, the network is available 24 a day 365 days a year.

There are also properties of networked cryptocurrencies related to their cryptographical nature.
These currencies do not require the users to trust a third party.
The trust relies on cryptographic primitives that have proven to be robust over the years.
The reliance on secret keys makes it possible to save the same money in more than one place at the same time.
Even if one of the copies of the secret keys is lost, the owner can still access the money if she has access to a second copy.
More sophisticated techniques can be used to achieve greater degrees of security. 
For example, the secret key can be divided in three parts in such a way that two out of the three parts are needed to access the money.
Secrets can also be kept in the brain, with no physical proof of the ownership of the money.

@@

The use of networked criptocurrencies evidences that flows of money are nothing more than flows of information.
Money transactions are simply communications among the involved parties.
The freedom to use different currencies is directly related to the freedom to communicate, which is popularly referred to as free speech.
The flows of money within a community are communications flows that strengthen the relation between the members of the community and, therefore, the community as a whole.

@@

The gift (or reputation) economy is the kind of economy we find on the Internet, in which the participants offer software, articles, music, pictures and all kind of digital goods for free.
This can be achieved using permissive licenses such as Creative Commons.
In the information age in which is obvious that money is just information and wealth is measured in knowledge instead of material goods, the progressive evolution towards a reputation economy is a natural step.

@@

Throughout human history, organizational models have changed, transformed and evolved.
As technology evolves, old models are been progressively replaced by others that more efficient in the context of the new technologies.
The invention of information processing and transmission tools multiplied the potential of community-centric models.
Initiatives embracing the community centric model have an advantage compared to those attaching to old hierarchical approaches.

It is natural that we find salient examples of community-centric success in the field of information and communication technologies.
The Internet itself is probably the best example of a community-centric endeavour.
There is a large number of people fascinated by and contributing to the Internet.
The decision-making forums are open and anyone is free to contribute expertise.
In a first level the users organize as networks, and then at another level those networks interconnect among them using peering agreements.

Another example of the successful application of the community-centric model is the free libre open source movement.
Thousands of programmers affiliate to and contribute to this movement working as a self-managed community.
The goods produced by this community often outperform those that have been developed using hierarchical models.
By making it possible for anyone to contribute, FLOSS projects have an advantage with respect to those that choose more restrictive models.

@@

In the case of open hardware, the fact that all schematics and manufacturing details are publicly available on the Internet makes it very economical for manufacturer to produce the goods.
The manufacturer can avoid all research and development costs and focus only in the fabrication.
As a consequence, open hardware devices can be found at very low prices in the market.
Lower prices translate to a greater community that can contribute to further improve the original product.

\section{Projects about Community Networks and Cryptocurrencies}
Community-centric innovation such as Community Networks and Networked Cryptocurrencies have attracted the attention of the research community.
In this section we detail three European projects that are related to community-centric innovation.

\subsection{Commons for Europe}
Commons for Europe is a project focused on stimulating the production of commons resources in the areas of applications and networks.
The branch of Commons for Europe related to application is called Code for Europe and invites cities to incorporate programmers to their teams.
These programmers bring community and collaborative dynamics to the host institution and write software that later can be shared with other cities.

The other branch of Commons for Europe is Bottom up Broadband.
This branch is focused on understanding and documenting the working principles of community networks.
